The role of standing variation in parallel adaptation

### 2014-09-13 standing-variation-paper.tex
minor edits

%\iflatexml  % no margin notes in latexml
% \newcommand{\marginnote}[1]{}
%\newcommand{\marginnote}[1]{{\it\color{red}(#1)}}
%\fi
\bibliographystyle{amnat}
%\bibliographystyle{plain}

@@

There is increasing number of examples of different populations within species often
adapt to shared environments via independent genetic changes. 
In some cases this convergent evolution leads to differing phenotypes
in response to shared selection pressures
\citep{} in other cases the degree of convergence is precise down the
same nucleutide change. Such convergent evoltuion within populations
has been seen for many
carefully studied phenotypes in a range of species, including drug resistance, pathogen resistance, insecticideresistance, and pigmentation. In many of these examples we 
need to be careful about whether they changes are truely convergent or
whether selection pressures differ subtly. 
We note that we treat the phrases convergent and parallel evolution
synonymously, as we are concerned with adaptation within a single
species that can occur via a different or shared genetic routes \citep{Arendt:08}.

The issue of convergent adaptation within species touches on a number
of important questions in evolutionary biology. These include the extent
to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, and whether
species should be regarded to cohesive units. Convergence evolution within
populations also effects our ability to detect adaptation from population genomic data as no one
allele sweeps fixation over the entire area affected by a selection
pressure \citep{softsweepsIII}.

Convergent evolution can occur even within a well mixed population
subject to a constant selection pressure, either through selection on
multiple mutations present as standing variation within population
before selection pressures switch \citep{Orr:01,softsweepsI} or due to
multiple adaptive alleles arising after selection pressures switch \citet{softsweepsII}. 
This previous work has shown that 
a primary determinant of the probability that multiple alleles contribute to adaptation
is the population scaled mutation rate \citep[see ][ for a review]{MesserPetrov}.

Geographically limited dispersal also increases the chance of
convergent evolution. For example when selection pressures are
geographically patchy this can increasing the probability of
convergent evolution as alleles are unable to spread through
intervening populations \citep{RalphCoop:14}. In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
where the selection pressure was constant across the entire environment, 
and that a single mutational change was sufficient to adapt populations to the novel environment. 
We assumed that there was no standing variation for the adaptive
allele, such that parallel mutation must be due to multiple mutations
after the environmental shift. In that setting we found that there was
a characteristic geographic scale, over which we would expect multiple
instances of our adaptive allele to have arisen in parallel. This
characteristic length could be expressed in terms of a simple compound
parameter determined by our parameters of interest. Here we extend
this spatial model include standing
variation present at mutation-selection balance before selection
pressures switch.

\subsection{The set-up}

@@

We assume that the species range
is a homogenous, one- or two-dimensional region. 
There are two selective classes -- the {\em neutral} type, and the {\em mutated} type.
We assume that different mutations are distinguishable --
either as selectively equivalent mutations, or by linked variation.
For a long time in the past, the mutated type has been at a selective disadvantage (long enough to be at selection-mutation equilibrium),
but at a certain time the selective regime changes, so that the mutated type has a selective advantage and quickly spreads to fixation.
After fixation, alleles are either descended
from families of mutants present as standing variation when the selective regime changed,
or from new mutants arising since that time.

For concreteness, suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type
(i.e.\ it produces on average $1-s_d$ times the number of offspring per generation),
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
where $s_b>0$ will usually be assumed to be small, and $0<s_d<1$.
We assume that diploid fitness is additive, or at least that the
important early dynamics are determined by the heterozygous fitness, with no reference to the fitness of the
  homozygote for the mutant alleles.
The number of offspring has finite variance, which to keep the formulas simple we assume to be equal to one.
(If the mutation is recessive, XXX more complicated things happen.)
As for the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the mean squared distance between parent and child is $\sigma^2$ (the {\em dispersal variance}).
The species occupies an area with mean density
$\rho$ haploid individuals (or chromosomes) per unit area.
\marginnote{XXX or do diploid? below is haploid.}

@@

\paragraph{Rates of origination of standing and new mutations}
We make use of the commonly used approximation that neglects competition between relatives,
treating the offspring of a new mutant that appears in an area not already occupied by the mutated type
as a branching process.
After $t=0$, the offspring of a single mutant form (approximately) a branching process with growth rate $s_b$,
so each new mutant establishes locally with probability $p_s \approx 2s_b$.
As in \cite{ralphcoop2010}, each new offspring has a very small probability of being a mutant and establishing locally,
so the collections of times and locations at which mutants appear and establish locally 
is well--approximated by a Poisson process in space and time.
This the rate of this Poisson process, 
the rate per unit area at which new mutants appear and establish locally after $t=0$ in areas not already occupied by the mutant type
is approximately $\lambda = 2 \mu \rho s_b$.

Before $t=0$, on the other hand, the mutation is deleterious, so the
genetic descendants of each new mutation are (with high probability)
doomed (doomed I say!) to extinction,
but may persist for some time.
Since the times and locations of mutants before $t=0$ also well-approximated by a Poisson process with rate $\mu \rho$,
the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
by the Mapping theorem, a Poisson process with rate we define to be $\lambda_0$.
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, we can then treat each such family as a single mutant,
equivalent to the mutants arising later.
(The approximation will be good if the log of the size of an extant families is small relative to the establishment time,
and the spatial distribution of the family is small relative to the spread between them.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
At time $t=0$, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s \approx 2s_b$ of establishing, approximately independently.
Therefore, the probability that some descendants of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the rate is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
Now for $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .
\end{align}
Note that this is closely related to the standard frequency of a
allele at mutation selection balance $\mu/s_d$ \citep{Haldane?}

@@

\paragraph{Geographic spread of alleles} 
Once an allele has become local established it can begin to spatially spread. 
We assume that the allele, having become established, quickly settles
down to spread spatially as a traveling wave of constant speed. The behavior of
this wave of advance of a beneficial allele was first described by 
\citet{fisher1937wave} and \citet*{KPP1937}. Under reasonably general
conditions the speed of advance
of this wave will be $v = \sigma \sqrt{2s}$. See
\citet{ralph2010parallel} for a fuller review of the Fisher-KPP as it
applies to this scenerio. Note that the speed of the wave will vary
with details of the space that individuals migrate across
\citep[e.g. see ][ for comparisons to migration on discrete
grids]{Slatkin-speed:76,SlatkinCharlesworth:78}, however, the velocity
will usually be primarily determined by the product $\sigma^2$ and $s$.

\paragraph{Putting it together.}
So we can put these ingredients together for a simple model of the
geographic spread of alleles. A cartoon example of this process is
shown in Figure \ref{fig:cartoon} Initially, when the selection
pressures change at $t=0$, a set of standing variants can start to
spread having escaped loss through drift. These variants are shown as
little lightning bolts, and occur at a density $\lambda_0$ across
space. They spread at velocity $v$ carving out cones in space-time. As these
alleles proceed in their geographic spread other new alleles can arise
and become established in parallel, these are shown as little
stars. These new mutations arise and become established at rate $\lambda$.

As we previously outlined in \citet{ralph2010parallel} this model of
geographic convergent evolution, when $\lambda_0=0$, is analogous to the \citet{kolmogorov-crystallization}
model of crystallization. In \citet{kolmogorov-crystallization} model nucleation points form at randomly at
a constant rate in time and space 
and initiate the radial growth of new crystals. The crystals at the
end point of this process form a random tessellation covering space
the properties of which have been extensively studied by \citet{moller92,moller95} and others \citep{bollobas-crystallization,gilbert-crystallization}. 
The generalized version of this process, for non-constant wave speeds
and imhomogeneous Poisson processes is termed the 
Kolmogorov-Johnson-Mehl-Avrami tessellation \citep[KJMA, see ][]{fanfoni-tomellini}.
Our combined process with both standing variation and new mutation is
a specially case of the KJMA tessellation, where the
spatial-temporally homogeneous
Poisson origination process of new mutations, is supplemented by a
single pulse of origination points at time zero with spatial density $\lambda_0$.

If we only allow standing variation then everything about the process
is relatively simple. All of the
points in space closer to a particular origination point will be
reached by that allele before any other
allele has a chance to spread there. Formally then the spatial patterning,
given a set of origination points, would be specified by the Voronoi
tessellation (i.e. the cells formed by assigning points to their
nearest origin). All of the properties of the patterning of space by alleles is
determined by the spatial locations of the initiation points, as
specified by the spatially homogeneous Poisson process, independent of
the rate of spatial spread. So that the properties
of the spatial tessellation with standing variation depends only on
$\lambda_0$, and this
model is commonly called the Poisson-Voronoi tesselation.

In our previous paper \citep{ralph2010parallel} we also provided results for non-constant wave
speeds, as can arise due to fat-tailed dispersal kernels. Here we
restict ourselves to constant wave speeds for simplicity but similar
calculations could be done by expansion of the results of the previous paper.

@@

\subsection{G6PD example}
We will describe a number of properties of our process, but we first
want to introduce a motivating example to provide some concrete
numbers to use for illustrative purposes.

Over the past ten thousand years malaria resistance alleles have arisen at a number
of genes and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to represent convergence, with
multiple changes in the same gene. 
For example, there are multiple
changes at the $\beta$-globin gene that confers malaria resistance,
and the sickle cell allele may plausibly have arisen by up to five distinct
changes at the same base pair within Africa
\citep{Flint:98,ralph2010parallel}.
Another particularly impressive case of convergent evolution in response to
selection pressures imposed by malaria are the numerous changes throughout the
X-linked G6PD gene, with upward of 50 polymorphic variants ($>1\%$ frequency) having so far been
described that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. These alleles are now
found at a combined frequency of $\sim$8 \% frequency in malaria endemic countries, rarely
exceeding$\sim$20 \% \citep{Howes-g6pd-preval}. Three malaria
resistance alleles at G6PD are particularly common and relatively well
studied: the $A-$ found in
much of sub-Saharan Africa; the Med allele found in the Mediterranean and middle
east; and the Mahidol allele found Myanmar and Thailand.
The $A-$ and Mahidol alleles have been shown
to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd,
Louicharoen-g6pd}. Haplotype-based analysis of genetic diversity surrounding
the $A-$, Med, and Mahidol suggest that they have spread in past few
thousand years \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other malaria resistance alleles. 
These population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, consistent with
the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the
$A-$ allele. 
Given such a strong pressure on these alleles they should have risen
quickly to fixation, so their presence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism. 
Hemizygous males and homozygous females suffer from G6PD deficiency,
leading to hemolytic anemia when exposed to a variety of compounds, notably
those present in fava beans.

The wave of advance works in the case of heterozygote advantage
\citep{aronson1975nonlinear}, with the selected allele spreading
locally to the equilibrium frequency (rather than fixation). As such
our framework is applicable here.

@@

The geographic area of Central and Eastern Asia with malaria is on the
order of $\sim 10$ million $km^2$. In that area there are at least $15$
common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). So the average width of
an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
The coding region of G6pd is 515 codons long, with $\sim$140
functional, distinct variants that cause reduced G6PD activity. Assuming a
mutation rate of $\approx 10^{-8}$ per base-pair per generation, we assume
$\mu \approx 1 \times 10^{-6}$ a generation. 
The dispersal and demographic parameters of humans in the past few
thousans years is unclear,
particularly as we are concerned with the effective population
density. 
We use $\rho=2$ and $0.2$ people per $km^2$ as two choices of our effective population density.
We use $\sigma=10,~50,$ and $100$ as our dispersal parameter. 
We note that obviously human migration has been
shaped both by local dispersal and larger-scale expansions \citep[see
][for a recent discussion]{}, so these parameters are for
illustrative purposes only.

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
\caption{ 
Map of G6pd-deficiency allele frequencies across Asia. The pie chart shows
  the frequency of G6pd-deficiency alleles. The size of the pie chart
  indicates the number of G6pd-deficient individuals sampled.
. Countries with
  endemic malaria are colored yellow. Figure taken from
  \citet{Howes-g6pd-variants}
  \url{http://www.malariajournal.com/content/12/1/418}. \gc{Not sure
    whether to include this.}
} \label{fig-G6PD-map}
\end{center}
\end{figure}

@@

%Selection coefficient (s) 	A- 0.044 (0.019–0.048)  Med	0.034
%(0.014–0.049)  ``The mean age of the A− allele in these runs was 6357 years, with a 95\% credibility interval extending from 3840 to 11,760 years. For the Med alleles, the mean age was 3330 years with a 95\% credibility interval of 1600 to 6640 years"
%\citet{tishkoff-g6pd}
%s 0.24 and t= 40 generations \citep{Slatkin-age-est}
%The estimated age is ~100 generations with an upper bound of <150
%generations A- 0.1 < s < 0.2 \citep{Saunders-g6pd}
%Mahidol487A mutation started to increase at ~1500 YBP, with a
%selection intensity of ~0.23 \citep{Louicharoen-g6pd}
%Show's effect in het and homozy females

@@

\subsection{Standing Variation -- Results} 
The quantity that is perhaps easiest to compute is the mean time until adaptation.
If fix ourselves at some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
Then, as shown in Figure XX,
$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
Since we assume these are a Poisson process, 
\[
    \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) ,
\]
and so we have
\begin{align}
    \E[\tau] % &= \int_0^\infty \P\{ \tau > t \} dt \\
        &= \int_0^\infty \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt.
\end{align}
For means of evaluting this integral, see Appendix \ref{apx:integrals}.

In Figure \ref{G6PD_chartimes}

@@

Furthermore, if we define $a_0$ to be the mean area occupied by a typical standing variant, 
then $a_0$ is given by the proportion of the range occupied by standing variants over the mean density of unique standing variants,
i.e.\ $a_0 = z_0 / \lambda_0$.
We can compare this to the corresponding quantity $a_+$ for a new variant through the formula
$a_0 / a_+ = z_0 / (1-z_0)$.

>>>>

%\iflatexml  % no margin notes in latexml
% \newcommand{\marginnote}[1]{}
%\newcommand{\marginnote}[1]{{\it\color{red}(#1)}}
%\fi
%% NEED amnat.bst IN REPO
% \bibliographystyle{amnat}
\bibliographystyle{plainnat}

@@

There are an increasing number of examples in which
different populations within species have
adapted to similar environments via independent genetic changes. 
In some cases this convergent evolution leads to differing phenotypes
in response to shared selection pressures
\citep{} in other cases the degree of convergence is precise down the
same nucleutide change. 
Such convergent evoltuion within populations has been seen for many
carefully studied phenotypes in a range of species, 
including drug resistance, pathogen resistance, insecticide resistance, and pigmentation. 
In many of these examples we 
need to be careful about whether the changes are truly convergent or
whether selection pressures differ subtly. 
We note that we treat the phrases ``convergent'' and ``parallel'' evolution
synonymously, as we are concerned with adaptation within a single
species that can occur via a different or shared genetic routes \citep[see][for more discussion]{Arendt:08}.

The issue of convergent adaptation within species touches on a number
of important questions in evolutionary biology. 
These include the extent to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, 
and to what degree should species be regarded to cohesive units. 
Convergent evolution within populations 
also affects our ability to detect adaptation from population genomic data,
since no single allele sweeps to fixation over the entire area affected by the selection pressure \citep{softsweepsIII}.

Convergent evolution can occur even within a well mixed population
subject to a constant selection pressure, 
either through selection on multiple mutations present as standing variation within population
before selection pressures switch \citep{Orr:01,softsweepsI},
or due to multiple adaptive alleles that arise after selection pressures switch \citet{softsweepsII}. 
The previous work has shown that 
a primary determinant of the probability that multiple alleles contribute to adaptation
is the population scaled mutation rate \citep[see ][ for a review]{MesserPetrov}.

\plr{should stick to either ``parallel'' or ``convergent'' perhaps}

Geographically limited dispersal also increases the chance of convergent evolution. 
For example, 
patchy geographically selection pressures can lead to much higher probability of parallel adaptation than uniform pressures,
since alleles are unable to spread through intervening populations \citep{RalphCoop:14}. 
In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
where the selection pressure was constant across the entire environment, 
and that a single mutational change was sufficient to adapt populations to the novel environment. 
We assumed that there was no standing variation for the adaptive allele, 
so that parallel mutation must be due to multiple mutations after the environmental shift. 
In that setting, 
we found that there was a characteristic geographic scale over which we would expect 
multiple instances of our adaptive allele to have arisen in parallel. 
This characteristic length could be expressed in terms of a simple compound
parameter determined by our parameters of interest. 
Here we extend this spatial model to include standing variation 
present at mutation-selection balance before the selection pressures switch.

%%%%%%%%% %%%%%%%%%%%% %%%%%%%
\subsection{The set-up}

@@

We assume that the species range
is a large, homogenous, one- or two-dimensional region. 
There are two selective classes -- the {\em neutral} type, and the {\em mutated} type.
We assume that separately arising mutations are distinguishable --
either as selectively equivalent mutations, or by linked variation.
We also suppose that 
the mutated type has been at a selective disadvantage 
for a sufficiently long enough time in the past to be at selection-mutation equilibrium,
but at a certain time the selective regime changes, so that the mutated type has a selective advantage and quickly spreads to fixation.
After fixation, alleles are either descended
from families of mutants present as standing variation when the selective regime changed,
or from new mutants arising since that time.

For concreteness, suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type
(i.e.\ it produces on average $1-s_d$ times the number of offspring per generation),
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
where $s_b>0$ will usually be assumed to be small, and $0<s_d<1$.
We assume that diploid fitness is additive, or at least that the
important early dynamics are determined by the heterozygous fitness, with no reference to the fitness of the
homozygote for the mutant alleles.
The number of offspring has finite variance, which to keep the formulas simple we assume to be equal to one.
(We leave aside the case where mutations are recessive, however.)
As for the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the mean squared distance between parent and child is $\sigma^2$.
The species occupies an area with mean density
$\rho$ haploid individuals (or chromosomes) per unit area.
\marginnote{XXX or do diploid? below is haploid.}

@@

\paragraph{Rates of origination of standing and new mutations}
We make use of the commonly used approximation that neglects competition between close relatives,
treating the offspring of a new mutant that appears in an area not already occupied by the mutated type
as a branching process.
After $t=0$, the offspring of a single mutant form (approximately) a branching process with growth rate $s_b$,
so each new mutant establishes locally with probability $p_s \approx 2s_b$.
As in \cite{ralphcoop2010}, each new offspring has a very small probability of being a mutant and establishing locally,
so the collections of times and locations at which mutants appear and establish locally 
is well--approximated by a Poisson process in space and time.
The rate of this Poisson process,
i.e.\ the rate per unit area at which new mutants appear and establish locally after $t=0$ in areas not already occupied by the mutant type,
is approximately $\lambda = 2 \mu \rho s_b$.

\plr{the following needs better explanation, maybe a figure?}
Before $t=0$, on the other hand, the mutation is deleterious.
We have assumed that $s_d$ is not too small
so that the genetic descendants of each new mutation are (with high probability) doomed to extinction,
but may persist for some time.
Since the times and locations of mutants before $t=0$ also well--approximated by a Poisson process with rate $\mu \rho$,
the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
by the Poisson Mapping Theorem, a Poisson process with rate we define to be $\lambda_0$.
% (In principle, more than one member of a mutant family may fix,
% but we assume these are close together enough that we may treat them as beginning from a single point.)
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, we can then treat each such family as a single mutant,
equivalent to the mutants arising later.
(The approximation will be good if the logarithm of the size of each extant family is small relative to the establishment time,
and the spatial distribution of each is small relative to the spread between the families.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
At time $t=0$, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s \approx 2s_b$ of establishing, approximately independently.
Therefore, the probability that some descendants of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the rate is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
Now for $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .
\end{align}
Note that this is closely related to the equilibrium frequency of a
allele at mutation selection balance $\mu/s_d$ \citep{Haldane?}

@@

\paragraph{Geographic spread of alleles} 
Once an allele has become locally established it can begin to spread across space. 
We assume that the allele, once established, 
quickly settles down to spread spatially as a traveling wave of constant speed. 
The behavior of this wave of advance of a beneficial allele was first described by 
\citet{fisher1937wave} and \citet*{KPP1937}. 
Under reasonably general conditions, the speed of advance of this wave is $v = \sigma \sqrt{2s}$. 
See \citet{ralph2010parallel} for a fuller review of these travelling waves.
Note that the speed of the wave will vary
with details of the space that individuals migrate across
\citep[e.g.\ see ][ for comparisons to migration on discrete grids]{Slatkin-speed:76,SlatkinCharlesworth:78}.
% However, the velocity will usually be primarily determined by the product of $\sigma^2$ and $s$.

\paragraph{Putting it together.}
Now, we can put these ingredients together for a simple model of the geographic spread of alleles. 
A cartoon example of this process is shown in Figure \ref{fig:cartoon}. 
Initially, when the selection pressures change at $t=0$, 
a set of standing variants can start to spread having escaped loss through drift. 
These variants are shown as little lightning bolts, and occur at a density $\lambda_0$ across space. 
They spread at velocity $v$, carving out cones in space-time. 
As these alleles proceed in their geographic spread, other new alleles can arise and become established in parallel, 
these are shown as little stars. 
These new mutations arise and become established at rate $\lambda$.

As we previously outlined in \citet{ralph2010parallel} this model of
geographic convergent evolution, when $\lambda_0=0$, is analogous to the a model of crystallization due to Kolmogorov \citet{kolmogorov-crystallization}.
In this model, nucleation points form at random at a constant rate in time and space 
and initiate the radial growth of new crystals. 
Eventually, the different orientations of crystals form a random tessellation of space,
hose properties have been extensively studied by \citet{moller92,moller95} and others \citep{bollobas-crystallization,gilbert-crystallization}. 
The generalized version of this process, for non-constant wave speeds
and inhomogeneous Poisson processes is termed the 
Kolmogorov-Johnson-Mehl-Avrami tessellation \citep[KJMA, see ][]{fanfoni-tomellini}.
Our combined process with both standing variation and new mutation is
a special case of the KJMA tessellation, 
where the spatial-temporally homogeneous Poisson origination process of new mutations, 
is supplemented by a single pulse of origination points at time zero with spatial density $\lambda_0$.
(For the purposes of analogy, we could imagine that before time $t=0$
the temperature is high enough that nucleation sites do not persist.)

If we only allow standing variation, then everything about the process is relatively simple. 
All of the points in space closer to a particular origination point will be
reached by that allele before any other allele has a chance to spread there. 
Such a tesselation of space is known as a Poisson-Voronoi tesselation \citep{poissonvoronoi}
(i.e.\ the cells formed by assigning points to the nearest point in a Poisson process).
The properties of the patterning of space by alleles is
determined by the spatial locations of the initiation points, 
as specified by the spatially homogeneous Poisson process, 
independent of the rate of spatial spread. 
Therefore, the properties of the spatial tessellation with standing variation depends only on $\lambda_0$.

% In our previous paper \citep{ralph2010parallel} we also provided results for non-constant wave
% speeds, as can arise due to fat-tailed dispersal kernels. 
% Here we restict ourselves to constant wave speeds for simplicity but similar
% calculations could be done by expansion of the results of the previous paper.

@@

%%%%%%%%% %%%%%%%%%%%

\subsection{G6PD example}

Below we describe a number of properties of our process, but we will first introduce 
a motivating example to provide some concrete
numbers to use for illustrative purposes.

Over the past ten thousand years, malaria resistance alleles have arisen at a number of genes 
and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to be examples of convergent adaptation, 
as different derived mutations in the same gene are seen in different individuals.
For example, 
a number of changes have been observed in the $\beta$-globin gene that confer malaria resistance;
and the sickle cell allele may plausibly have arisen by up to five independent occurrences
of the same base pair mutatoin at different locations within Africa
\citep{Flint:98,ralph2010parallel}.
Another particularly impressive case of convergent evolution in response to
selection pressures imposed by malaria are the numerous changes throughout the X-linked G6PD gene, 
with upward of 50 polymorphic variants ($>1\%$ local frequency) having so far been described 
that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
These alleles are now found at a combined frequency of around 8\% frequency in malaria endemic areas,
rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
Three G6PD deficiency alleles that also confer resistance to malaria 
are particularly common and relatively well studied: 
the $A-$ allele found in much of sub-Saharan Africa; 
the Med allele found in the Mediterranean and Middle East; 
and the Mahidol allele found Myanmar and Thailand.
The $A-$ and Mahidol alleles have been shown
to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd, Louicharoen-g6pd}. 
Haplotype-based analysis of genetic diversity surrounding
the $A-$, Med, and Mahidol suggest that they have spread over the past 
few thousand years \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other known malaria resistance alleles. 
Population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the $A-$ allele. 
\plr{what are Ruwende's results? feels a bit vague here}
Given such a strong pressure on these alleles they should have risen quickly to fixation, 
so their presence at intermediate frequency over a broad geographic area
makes them good candidates for recently balanced polymorphisms. 
Hemizygous males and homozygous females suffer from G6PD deficiency,
leading to hemolytic anemia when exposed to a variety of compounds, 
notably those present in fava beans.

Note that although originally the ``wave of advance'' was presented as the spread of an unconditionally favorable allele
\citep{fisher1936,KPP},
the same model holds in the case of heterozygote advantage \citep{aronson1975nonlinear}, 
with the selected allele spreading locally to the equilibrium frequency (rather than fixation),
with speed determined by the advantage of heterozygotes when rare.

@@

The geographic area of Central and Eastern Asia with malaria is on the
order of $\sim 10$ million $km^2$. In that area there are at least $15$
common variants \citep{Howes-g6pd-variants} (see Figure \ref{fig-G6PD-map}). 
Therefore, the average width of an area occupied by an allele is $\sqrt{10^7/15} \approx  800$km. 
The coding region of G6pd is 515 codons long, 
and around 140 distinct deficiency alleles have been observed. 
Assuming a mutation rate of $\approx 10^{-8}$ per base-pair per generation, 
we can take $\mu \approx 1 \times 10^{-6}$ per generation. 
The dispersal and demographic parameters of humans in the past few thousans years is unclear,
particularly as we are concerned with the ``effective'' population density
(i.e.\ population density divided by variance in offspring number).
We therefore will use two reasonable values for the effective population density: $\rho=2$ and $0.2$ people per $km^2$,
and three values for the dispersal distance: $\sigma=10$, $50$ and $100$.
Note that obviously human migration has been shaped both by local dispersal and larger-scale expansions 
\citep[see ][for a recent discussion]{}, 
so these parameters only provide a rough view of the process.

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
\caption{ 
Map of G6pd-deficiency allele frequencies across Asia. 
The pie chart shows the frequency of G6pd-deficiency alleles. 
The size of the pie chart indicates the number of G6pd-deficient individuals sampled.
Countries with endemic malaria are colored yellow. 
Figure taken from \citet{Howes-g6pd-variants}
\url{http://www.malariajournal.com/content/12/1/418}. 
\gc{Not sure whether to include this.}
\plr{I like it.}
} \label{fig-G6PD-map}
\end{center}
\end{figure}

@@

%Selection coefficient (s) 	A- 0.044 (0.019–0.048)  Med	0.034
%(0.014–0.049)  ``The mean age of the A- allele in these runs was 6357 years, with a 95\% credibility interval extending from 3840 to 11,760 years. For the Med alleles, the mean age was 3330 years with a 95\% credibility interval of 1600 to 6640 years"
%\citet{tishkoff-g6pd}
%s 0.24 and t= 40 generations \citep{Slatkin-age-est}
%The estimated age is ~100 generations with an upper bound of <150
%generations A- 0.1 < s < 0.2 \citep{Saunders-g6pd}
%Mahidol487A mutation started to increase at ~1500 YBP, with a
%selection intensity of ~0.23 \citep{Louicharoen-g6pd}
%Shows effect in het and homozy females

%%%%%%%%%%%% %%%%%%%%%%%%%%%

@@

%%%%%%%%%%% %%%%%%%%%%%%% %%%%%%%%%
\subsection{Time to adaptation}

The quantity that is perhaps easiest to compute is the mean time until adaptation.
If fix ourselves at some geographic location, and let $\tau\ge0$ be the time at which the point is reached by the advantageous mutations.
Then, as can be seen from Figure~\ref{fig:cartoon},
$\tau > t$ if and only if the cone with point at $(x,t)$ and slope $v$ extending back to $t=0$ is empty of successful mutations.
Since we assume these are a Poisson process, 
\[
    \P\{ \tau > t \} = \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) ,
\]
and since $\E[\tau] = \int_0^\infty \P\{ \tau > t \}$,
\begin{align}
    \E[\tau] % &= \int_0^\infty \P\{ \tau > t \} dt \\
        &= \int_0^\infty \exp\left( - \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt.
\end{align}
For means of evaluting this integral, see Appendix \ref{apx:integrals}.
In Figure \ref{G6PD_chartimes} we show the mean time until adaptation for various values of the parameters.

@@

Furthermore, if we define $a_0$ to be the mean area occupied by a typical standing variant, 
then $a_0$ is given by the proportion of the range occupied by standing variants divided by the mean density of unique standing variants,
i.e.\ $a_0 = z_0 / \lambda_0$.
We can solve for $a_+$, the corresponding mean area occupied by a given new variant, 
using the formula $a_0 / a_+ = z_0 / (1-z_0)$.
